Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-10-16 paper.qmd
Added PaCMAP model for two C-shaped data

```{r}
#| label: best-fit-umap-tsne
#| fig-cap: "Best fit for two C-shaped clusters data using UMAP (a1-a4) ($a_1$ = 0.09) and tSNE (b1-b4) ($a_1$ = 0.16) with the same number of non-empty bins ($m$ = 39). Both models show twist in $7\\text{-}D$ (a3, a4, b3, b4). Also, both NLDR methods are unable to accurately capture the width of the curvilinear clusters."
#| fig-width: 10
#| fig-height: 20
#| out-height: 90%

nldr_two_curvy_original2 + tsne_two_c_shaped_clust_plt +
  trimesh_two_curvy_removed1_with_data + trimesh_two_curvy_removed1_tsne_with_data +
  two_curvy_proj_umap_model2 +
  two_curvy_proj_tsne_model +
  two_curvy_proj_umap_all_model +
  two_curvy_proj_tsne_all_model +
  plot_layout(ncol = 2)
```

<!-- @fig-two_curvy-sc-best shows one of the best fits for S-curve data. TriMAP with $15$ nearest neighbors for forming the nearest neighbor triplet (`n_inliers`), $3$ outliers for forming the nearest neighbor triplets (`n_outliers`), $3$ random triplets per point (`n_random`) is used as the NLDR method. The model is generated with $103$ bins, which have a width of $0.154$ and a height of $0.133$. Edges with lengths greater than $0.266$ are removed.  -->

<!-- The appropriate model for the S-curve using UMAP (n_neighbors: $15$) was created with $420$ bins. It was determined to be the best model after removing edges with a length greater than $0.134$. The decision was made considering that the model accurately captures the geometry of the S-curve. But better in some places only (@fig-scurve_sc_best). -->

<!-- In contrast, a poor model for the S-curve with PHATE was created using only $190$ bins and removing edges with a length greater than $0.3$. This model is inadequate as it has a higher MSE compared to the appropriate model. Visually, the model squeezed to the middle of the geometry of the S-curve (@fig-scurve-sc-bad). -->

<!-- XXX BEST FIT FOR S-CURVE MIGHT BE A DIFFERENT NLDR.  -->
<!-- SHOW THE SEVERAL SUB-OPTIMAL FITS. -->

<!--langevitour with best pD model for S-curve--> 
<!-- ::: {#fig-scurve-sc-best layout-ncol="3" fig-pos="H"} -->
<!-- ![](figures/scurve/sc_best_1.png) -->

<!-- ![](figures/scurve/sc_best_2.png) -->

<!-- ![](figures/scurve/sc_best_3.png) -->

<!-- Screen shots of the **langevitour** of the S-curve, shows the model-in-data space, a video of the tour animation is available at (<https://youtu.be/4RFNENX_fXk>). MSE with $0.035$. -->
<!-- ::: -->

<!--langevitour with bad pD model for S-curve--> 
<!-- ::: {#fig-scurve-sc-bad layout-ncol="3" fig-pos="H"} -->
<!-- ![](figures/scurve/sc_bad_1.png) -->

<!-- ![](figures/scurve/sc_bad_2.png) -->

<!-- ![](figures/scurve/sc_bad_3.png) -->

<!-- Screen shots of the **langevitour** of the S-curve, shows the model-in-data space, a video of the tour animation is available at (<https://youtu.be/DHPvhkZ0lUM>). MSE with $0.090$. -->
<!-- ::: -->

<!------>
<!--add more details regarding to the how the change happening with different number of bins-->

<!--best model-->
<!--need to mention why it is the best-->

<!--bad model-->
<!--need to mention why it is the bad-->
<!--Not represented the whole data structure because there are some discontinuity in the S-curve-->
<!-- Why? bad

<!-- add MSE values for the screenshot figure names-->

>>>>

```{r}
#| label: best-fit-pacmap-two-curvy

two_curvy_pacmap_scaled_obj <- gen_scaled_data(
  data = pacmap_two_c_shaped_clust)

pacmap_two_curvy_scaled <- two_curvy_pacmap_scaled_obj$scaled_nldr |>
  mutate(ID = row_number())
lim1 <- two_curvy_pacmap_scaled_obj$lim1
lim2 <- two_curvy_pacmap_scaled_obj$lim2
r2_pacmap <- diff(lim2)/diff(lim1)

pacmap_two_c_shaped_clust_plt <- pacmap_two_curvy_scaled |>
  ggplot(aes(x = PaCMAP1,
             y = PaCMAP2))+
  geom_point(alpha=0.1, size=1, colour='#e41a1c') +
  interior_annotation("b1", sc_ltr_pos) +
  theme(aspect.ratio = 1)

## hexagon binning to have regular hexagons
hb_obj_pacmap_two_curvy2 <- hex_binning(
  data = pacmap_two_curvy_scaled, 
  bin1 = 8, 
  r2 = r2_pacmap)

a1_2_pacmap <- calc_bins_y(
  bin1 = 8, 
  r2 = r2_pacmap
)$a1

## Data set with all centroids
all_centroids_df4 <- hb_obj_pacmap_two_curvy2$centroids

## Generate all coordinates of hexagons
hex_grid4 <- hb_obj_pacmap_two_curvy2$hex_poly

## To obtain the standardise counts within hexbins
counts_df4 <- hb_obj_pacmap_two_curvy2$std_cts
df_bin_centroids4 <- extract_hexbin_centroids(
  centroids_df = all_centroids_df4, 
  counts_df = counts_df4) |>
  filter(drop_empty == FALSE) |>
  mutate(b1 = "b1 = 11")

pacmap_data_two_curvy2_with_hb_id <- hb_obj_pacmap_two_curvy2$data_hb_id
df_all_two_curvy4 <- dplyr::bind_cols(training_data_two_curvy, pacmap_data_two_curvy2_with_hb_id)
df_bin_two_curvy4 <- avg_highd_data(data = df_all_two_curvy4, col_start = "x")

trimesh_two_curvy_removed1_pacmap_with_data <- ggplot() +
  geom_segment(data = tr_from_to_df_two_curvy4,
               aes(
                 x = x_from,
                 y = y_from,
                 xend = x_to,
                 yend = y_to),
               colour = "#000000") +
  geom_point(data = pacmap_two_curvy_scaled,
             aes(
               x = PaCMAP1,
               y = PaCMAP2
             ),
             color = "#a65628",
             alpha = 0.5,
             size = 0.5) +
  coord_equal() +
  interior_annotation("b2", sc_ltr_pos)

```{r}
#| label: best-fit-pacmap-two-curvy-proj

two_curvy_proj_pacmap_model <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#000000") + #31a354
  geom_point(
    size = 0.5,
    alpha = 0.5,
    color = "#a65628") +
  geom_segment(
    data=axes,
    aes(x=x1, y=y1, xend=x2, yend=y2),
    colour="grey70") +
  geom_text(
    data=axes,
    aes(x=x2, y=y2), 
    label=rownames(axes),
    colour="grey50",
    size = 4) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-0.126, 0.125)) + 
  ylim(c(-0.126, 0.125)) +
  interior_annotation("b3", sc_ltr_pos) +
  theme(
    legend.position = "none"
  )

two_curvy_proj_pacmap_all_model <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_segment(
    data = true_model_df_proj, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#2b8cbe") +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#000000") +
  geom_point(
    size = 0.5,
    alpha = 0.5,
    color = "#a65628") +
  geom_segment(
    data=axes,
    aes(x=x1, y=y1, xend=x2, yend=y2),
    colour="grey70") +
  geom_text(
    data=axes,
    aes(x=x2, y=y2), 
    label=rownames(axes),
    colour="grey50",
    size = 4) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-0.126, 0.125)) + 
  ylim(c(-0.126, 0.125)) +
  interior_annotation("b4", sc_ltr_pos) +
  theme(
    legend.position = "none"
  )
```

```{r}
#| label: best-fit-umap-tsne-pacamp
#| fig-cap: "Best fit for two C-shaped clusters data using UMAP (a1-a4) ($a_1$ = 0.09) and tSNE (b1-b4) ($a_1$ = 0.16) with the same number of non-empty bins ($m$ = 39). Both models show twist in $7\\text{-}D$ (a3, a4, b3, b4). Also, both NLDR methods are unable to accurately capture the width of the curvilinear clusters."
#| fig-width: 15
#| fig-height: 20
#| out-height: 90%

nldr_two_curvy_original2 + tsne_two_c_shaped_clust_plt +
  pacmap_two_c_shaped_clust_plt +
  trimesh_two_curvy_removed1_with_data + trimesh_two_curvy_removed1_tsne_with_data +
trimesh_two_curvy_removed1_pacmap_with_data +
  two_curvy_proj_umap_model2 +
  two_curvy_proj_tsne_model +
  two_curvy_proj_pacmap_model +
  two_curvy_proj_umap_all_model +
  two_curvy_proj_tsne_all_model +
  two_curvy_proj_pacmap_all_model +
  plot_layout(ncol = 3)
```
